Learning with misspecified models: Overconfidence and Stereotypes

### 2023-10-09 Paper/draft.Rmd
lab slides

@@

I recruited `r print(unique(updates$code))` undergraduate subjects from the CESS lab at NYU who participated in an in-person experiment. 
Sessions lasted approximately 
45 minutes and subjects earned an average payment of $\$22.6$. The experiment was programmed using oTree [@chen2016otree].

@@

In order to analyze the learning about the state, I look at the share of optimal choices that are made at each round.
I find that although subjects seem to be improving in their choices overall, the last choice coincided 
with the true state only in 52% of the choices in the last round of each task. This is statistically 
greater than the initial share of optimal choices of 30% ($p<0.01$). However, it is still far from 
complete learning. It is also important to note that learning is similar across treatments. The 
share of choices that were consistent with the true state for each round is reported in Figure \ref{fig:optimal-choices-by-round}.

@@

A closer look at whether people learned or not reveals that there is a good amount of heterogeneity in the
sample. Figure \ref{fig:learning-by-groups} shows the share of optimal choices by round for 
subjects that chose an effort that matched the state in 3 out of the last 4 rounds. It also shows the 
share of optimal choices for subjects who chose an effort that matched the state in fewer than 3 out of 
the last 4 rounds. I label the former as learners and the latter as non-learners, with learners making up
`r round(sum(updates$learner)/nrow(updates), 2)*100`% of the sample.

@@

I look at the subjects who are updating in a task where the combination of $\theta$ and $\omega$ is such that
there is a self-defeating equilibrium. I then look at how often

@@

@@

>>>>

\begin{table}[htbp]
\centering

@@

I recruited `r print(unique(updates$code))` undergraduate subjects from the CESS lab at NYU who participated in an in-person experiment. 
Sessions lasted approximately 
45 minutes and subjects earned an average payment of $\$22.6$. The experiment was programmed using oTree [@otree].

@@

In order to analyze the learning about the state, I look at the share of optimal choices that are made at each round.
I find that although subjects seem to be improving in their choices overall, the last choice coincided 
with the true state only in 52% of the choices in the last round of each task. This is statistically 
greater than the initial share of optimal choices ($p<0.01$). However, it is still far from 
complete learning. It is also important to note that learning is similar across treatments. The 
share of choices that were consistent with the true state for each round is reported in Figure \ref{fig:optimal-choices-by-round}.

@@

A closer look at whether people learned or not reveals that there is a good amount of heterogeneity in the
sample. Figure \ref{fig:learning-by-groups} shows the share of optimal choices by round for 
subjects that chose an effort that matched the state in 3 out of the last 4 rounds. It also shows the 
share of optimal choices for subjects who chose an effort that matched the state in fewer than 3 out of 
the last 4 rounds. I label the former as learners and the latter as non-learners, with learners making up
`r round(sum(updates$learned)/nrow(updates), 2)*100`% of the sample.

@@

Overall, it does not seem to be the presence of traps that is preventing learning. Figure \ref{fig:traps}
shows that the share of subjects who learned in cases where there was a trap is larger than the share of
subjects who learned in cases where there was no trap. This is true for both treatments. This indicates
that even in the presence of traps, some people learn.

![Share of subjects who learned in cases where there was a trap and in cases where there was no trap](../figures/learning_trap_presence.png){#fig:traps}

A closer look at the behavior of only the cases in which there were traps reveals that among those who did not
learn, the choices that support stable beliefs were chosen much more often than among subjects who learned. 
Figure \ref{fig:traps-choice} shows the share of choices that were consistent with a stable belief for the
learners and the non-learners. The difference is statistically significant ($p<0.01$). Thus, although the 
presence of self-defeating equilibria is not the main deterrent to learning, it does seem to be a factor
in the behavior of those who do not learn.

![Share of choices that were consistent with a stable belief for the learners and the non-learners](../figures/trap_choice.png){#fig:traps-choice}

```{r, echo=FALSE, message=FALSE, warning=FALSE}
share_trapped <- round(sum(updates$trapped) / length(updates$trapped),2)
```

When looking a the behavior in the last 4 rounds of each task, only `r print(share_trapped)*100`% 
of the subjects  
were trapped\footnote{I consider a subject to be trapped if 3 out of the last 4 choices were consistent 
with the self-defeating equilibrium} in a self-defeating equilibrium. These subjects could have been
dogmatic, switchers for whom the likelihood did not grow fast enough, or Bayesian agents who fell into 
a self-defeating equilibrium. The rest of the subjects were either updating their beliefs with a bias
or were switchers.

So far I have accounted for `r share_trapped*100`% subjects as being trapped, and 
`r round(sum(updates$learned)/nrow(updates), 2)*100`%
having learned the state correctly. From the remaining `r (1 - share_trapped - round(sum(updates$learned)/nrow(updates), 2))*100`%, 
`r round(sum(updates$categorized=='neither' & updates$trap==1)/sum(updates$categorized=='neither'), 2)*100`% were prone to traps
but did not fall into them, therefore they could not have been dogmatic modelers. Alternative explanations
are that they were switchers who considered an incorrect alternative model, or that they were updating with a bias so large that their effort
choices were far from optimal.

In the following section, I look at indications of biased updating among all the groups identified so far.

@@

One of the key features of the self-attribution bias is that the agent will react differently to good news and bad news.
In particular, they will overreact to bad news and underreact to good news. I define good news to be any sample of outcomes
for which the realized number of successes is greater than the average number of successes so far. Similarly, I define bad news to be
any sample of outcomes for which the realized number of successes is smaller than the average number of successes so far.
In addition, I calculate the difference between the realized number of successes and the average number of successes so far and
call it the news differential.
Figure \ref{fig:good-bad-news} shows the correlation between change in effort and the news differential, separately for 
good and bad news.

The overall pattern is that bad news are associated with a decrease in effort, and a 
more negative news differential is associated with a larger decrease in effort. On the contrary, good news are associated with 
a small decrease in effort. As mentioned in the predictions section, this is consistent with the self-attribution bias as well
as with a change of paradigm. It is not consistent with the dogmatic modeler or the Bayesian. Because 
I do not directly elicit beliefs, it is not possible to distinguish between the self-attribution bias and the change of 
paradigm with this data alone.
Nevertheless, the fact that about a third of the subjects started with a correctly specified belief and ended up overconfident
is consistent only with the self-attribution bias. Therefore, I conclude that the self-attribution bias is the most likely
explanation for the behavior of the subjects in the experiment.

![Change in effort by news differential](../figures/effort_change_news.png){#fig:good-bad-news}

A regression analysis of this same data reveals that indeed, the slopes are different for the two types 
of news and that the difference is statistically significant ($p<0.01$). The regression results are reported in Table \ref{tab:regression}.

```{r, echo=FALSE, message=FALSE, warning=FALSE }
reg <- lm(effort_change ~ good_news + news_diff + news_diff:good_news, data=updates[round_number>1,])
robust_se <- sqrt(diag(vcovHC(reg)))

# Attach the robust standard errors to the coefficient estimates
coefs <- coef(reg)
coefs <- cbind(coefs, robust_se)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# run the regression only for treatment  = ego
reg_ego <- lm(effort_change ~ good_news + news_diff + news_diff:good_news, data=updates[round_number>1 & treatment == 'ego',])

# Attach the robust standard errors to the coefficient estimates
coefs_ego <- coef(reg_ego)
coefs_ego <- cbind(coefs_ego, robust_se)

# run the regression only for treatment  = stereotype
reg_stereotype <- lm(effort_change ~ good_news + news_diff + news_diff:good_news, data=updates[round_number>1 & treatment == 'stereotype',])

# Attach the robust standard errors to the coefficient estimates
coefs_stereotype <- coef(reg_stereotype)
coefs_stereotype <- cbind(coefs_stereotype, robust_se)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# report reg, reg_ego, reg_stereotype in different columns using stargazer

stargazer(reg, reg_ego, reg_stereotype, 
title='Regression of effort change on news differential', 
          dep.var.labels='Change in effort', 
          covariate.labels=c('Good News', 'News differential', 'News Differential * Good News'),
          column.labels=c('All', 'Ego-relevant', 'Stereotype'), 
          omit.stat=c('f', 'ser'), 
          digits=2, 
          label='tab:regression', 
          header=FALSE, 
          se = list(coefs[, 2], coefs_ego[, 2], coefs_stereotype[, 2]))
```

# Structural estimation

## Model fit

@@

I  designed a setting in which I can distinguish between the forces of dogmatism, paradigm shifts, and optimal expectations.
Through a laboratory experiment, I collected data on behavior and infer beliefs from it. I find that 
the behavior of the subjects is consistent with the self-attribution bias. In particular, I find that 
subjects decrease their effort in response to bad news as well as to bad news, although the latter decrease 
is smaller. This is consistent with the self-attribution bias as well as with a change of paradigm. However,
I also observe that about a third of the subjects started with a correctly specified belief and ended up
overconfident. This is consistent only with the self-attribution bias. Therefore, I conclude that the
model that best explains the data is that of optimal expectations.

I have small or no effects throughout on the ego-relevance of the type. Although the motivation behind the 
theories does not directly apply to the case of stereotypes, the lack of stark differences suggest that 
biases arise due to an incorrect updating procedure and not necessarily due to the self-serving motivation 
behind the bias. The lack of differences in the behavior of the subjects across treatments could also 
be due to self-censoring of the subjects. In particular, subjects might be initially reluctant to 
express their true belief about a particular group of people. In addition, as they learn and realize 
that their beliefs might be incorrect, they might overreact to signals to try to compensate for the initial 
bias. Such an updating procedure would be consistent with the self-attribution bias for the case of sterotypes.
I remain agnostic about whether that is truly the case or whether the lack of differences is due to other 
factors.

Another bias that has been studied and suggested in the literature is the confirmation bias. In particular,
it has been suggested that subjects who are overestimate $\theta$ in their prior, will over-attribute 
successes to a high type and under-attribute them to a low type. Symmetrically, subjects who initially 
underestimate $\theta$ will over-attribute failures to a low type and under-attribute them to a high type.
I do not find evidence of such a bias in the data. This is consistent with what @Mobius2022 find in their
experiment.

I further investigate possible heterogeneity in the underlying models being used by estimating the 
structural parameters of the models.

### 2023-10-09 Paper/draft.tex
lab slides

\title{Learning with misspecified models: Overconfidence and
Stereotypes}
\author{Jimena Galindo}
\date{October 07, 2023}

@@

I recruited undergraduate subjects from the CESS lab at NYU who
participated in an in-person experiment. Sessions lasted approximately
45 minutes and subjects earned an average payment of \(\$22.6\). The
experiment was programmed using oTree \citep{chen2016otree}.

@@

In order to analyze the learning about the state, I look at the share of
optimal choices that are made at each round. I find that although
subjects seem to be improving in their choices overall, the last choice
coincided with the true state only in 52\% of the choices in the last
round of each task. This is statistically greater than the initial share
of optimal choices of 30\% (\(p<0.01\)). However, it is still far from
complete learning. It is also important to note that learning is similar
across treatments. The share of choices that were consistent with the
true state for each round is reported in Figure
\ref{fig:optimal-choices-by-round}.

@@

A closer look at whether people learned or not reveals that there is a
good amount of heterogeneity in the sample. Figure
\ref{fig:learning-by-groups} shows the share of optimal choices by round
for subjects that chose an effort that matched the state in 3 out of the
last 4 rounds. It also shows the share of optimal choices for subjects
who chose an effort that matched the state in fewer than 3 out of the
last 4 rounds. I label the former as learners and the latter as
non-learners, with learners making up 0\% of the sample.

@@

This transition matrix

@@

@@

>>>>

\title{Learning with misspecified models: Overconfidence and
Stereotypes}
\author{Jimena Galindo}
\date{October 08, 2023}

@@

I recruited undergraduate subjects from the CESS lab at NYU who
participated in an in-person experiment. Sessions lasted approximately
45 minutes and subjects earned an average payment of \(\$22.6\). The
experiment was programmed using oTree \citep{otree}.

@@

In order to analyze the learning about the state, I look at the share of
optimal choices that are made at each round. I find that although
subjects seem to be improving in their choices overall, the last choice
coincided with the true state only in 52\% of the choices in the last
round of each task. This is statistically greater than the initial share
of optimal choices (\(p<0.01\)). However, it is still far from complete
learning. It is also important to note that learning is similar across
treatments. The share of choices that were consistent with the true
state for each round is reported in Figure
\ref{fig:optimal-choices-by-round}.

@@

A closer look at whether people learned or not reveals that there is a
good amount of heterogeneity in the sample. Figure
\ref{fig:learning-by-groups} shows the share of optimal choices by round
for subjects that chose an effort that matched the state in 3 out of the
last 4 rounds. It also shows the share of optimal choices for subjects
who chose an effort that matched the state in fewer than 3 out of the
last 4 rounds. I label the former as learners and the latter as
non-learners, with learners making up 39\% of the sample.

@@

The transition matrix points towards the presence of self-attribution
bias in the data. However, it is still possible that there is
heterogeneity in the updating procedure across subjects. In the next
section I will look at the role of self-defeating equilibria in
preventing learning. I find that the presence of self-defeating
equilibria does not seem to be the main deterrent to learning.

\hypertarget{learning-traps}{%
\subsection{Learning traps}\label{learning-traps}}

As described above, the presence of stable misspecified beliefs act as
learning traps. If an agent falls into one of these traps, bayesian
updating will lead them to choose the same effort forever. These
self-defeating equilibria allow me to distinguish between the dogmatic
modeler and the switcher. If the agent falls into a trap and stays there
forever, they are more likely to be a dogmatic modeler. If they escape
the trap, they are more likely to be a switcher or to be updating with
some attribution bias.

Overall, it does not seem to be the presence of traps that is preventing
learning. Figure \ref{fig:traps} shows that the share of subjects who
learned in cases where there was a trap is larger than the share of
subjects who learned in cases where there was no trap. This is true for
both treatments. This indicates that even in the presence of traps, some
people learn.

\begin{figure}
\hypertarget{fig:traps}{%
\centering
\includegraphics{../figures/learning_trap_presence.png}
\caption{Share of subjects who learned in cases where there was a trap
and in cases where there was no trap}\label{fig:traps}
}
\end{figure}

A closer look at the behavior of only the cases in which there were
traps reveals that among those who did not learn, the choices that
support stable beliefs were chosen much more often than among subjects
who learned. Figure \ref{fig:traps-choice} shows the share of choices
that were consistent with a stable belief for the learners and the
non-learners. The difference is statistically significant (\(p<0.01\)).
Thus, although the presence of self-defeating equilibria is not the main
deterrent to learning, it does seem to be a factor in the behavior of
those who do not learn.

\begin{figure}
\hypertarget{fig:traps-choice}{%
\centering
\includegraphics{../figures/trap_choice.png}
\caption{Share of choices that were consistent with a stable belief for
the learners and the non-learners}\label{fig:traps-choice}
}
\end{figure}

When looking a the behavior in the last 4 rounds of each task, only 16\%
of the subjects\\
were
trapped\footnote{I consider a subject to be trapped if 3 out of the last 4 choices were consistent 
with the self-defeating equilibrium} in a self-defeating equilibrium.
These subjects could have been dogmatic, switchers for whom the
likelihood did not grow fast enough, or Bayesian agents who fell into a
self-defeating equilibrium. The rest of the subjects were either
updating their beliefs with a bias or were switchers.

So far I have accounted for 16\% subjects as being trapped, and 39\%
having learned the state correctly. From the remaining 45\%, 60\% were
prone to traps but did not fall into them, therefore they could not have
been dogmatic modelers. Alternative explanations are that they were
switchers who considered an incorrect alternative model, or that they
were updating with a bias so large that their effort choices were far
from optimal.

In the following section, I look at indications of biased updating among
all the groups identified so far.

@@

One of the key features of the self-attribution bias is that the agent
will react differently to good news and bad news. In particular, they
will overreact to bad news and underreact to good news. I define good
news to be any sample of outcomes for which the realized number of
successes is greater than the average number of successes so far.
Similarly, I define bad news to be any sample of outcomes for which the
realized number of successes is smaller than the average number of
successes so far. In addition, I calculate the difference between the
realized number of successes and the average number of successes so far
and call it the news differential. Figure \ref{fig:good-bad-news} shows
the correlation between change in effort and the news differential,
separately for good and bad news.

The overall pattern is that bad news are associated with a decrease in
effort, and a more negative news differential is associated with a
larger decrease in effort. On the contrary, good news are associated
with a small decrease in effort. As mentioned in the predictions
section, this is consistent with the self-attribution bias as well as
with a change of paradigm. It is not consistent with the dogmatic
modeler or the Bayesian. Because I do not directly elicit beliefs, it is
not possible to distinguish between the self-attribution bias and the
change of paradigm with this data alone. Nevertheless, the fact that
about a third of the subjects started with a correctly specified belief
and ended up overconfident is consistent only with the self-attribution
bias. Therefore, I conclude that the self-attribution bias is the most
likely explanation for the behavior of the subjects in the experiment.

\begin{figure}
\hypertarget{fig:good-bad-news}{%
\centering
\includegraphics{../figures/effort_change_news.png}
\caption{Change in effort by news differential}\label{fig:good-bad-news}
}
\end{figure}

A regression analysis of this same data reveals that indeed, the slopes
are different for the two types of news and that the difference is
statistically significant (\(p<0.01\)). The regression results are
reported in Table \ref{tab:regression}.

\begin{table}[!htbp] \centering 
  \caption{Regression of effort change on news differential} 
  \label{tab:regression} 
\begin{tabular}{@{\extracolsep{5pt}}lccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{3}{c}{\textit{Dependent variable:}} \\ 
\cline{2-4} 
\\[-1.8ex] & \multicolumn{3}{c}{Change in effort} \\ 
 & All & Ego-relevant & Stereotype \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 Good News & $-$0.12$^{**}$ & $-$0.16$^{***}$ & $-$0.05 \\ 
  & (0.05) & (0.05) & (0.05) \\ 
  & & & \\ 
 News differential & 0.22$^{***}$ & 0.22$^{***}$ & 0.21$^{***}$ \\ 
  & (0.02) & (0.02) & (0.02) \\ 
  & & & \\ 
 News Differential * Good News & $-$0.27$^{***}$ & $-$0.25$^{***}$ & $-$0.29$^{***}$ \\ 
  & (0.02) & (0.02) & (0.02) \\ 
  & & & \\ 
 Constant & 0.31$^{***}$ & 0.31$^{***}$ & 0.30$^{***}$ \\ 
  & (0.04) & (0.04) & (0.04) \\ 
  & & & \\ 
\hline \\[-1.8ex] 
Observations & 4,680 & 2,700 & 1,980 \\ 
R$^{2}$ & 0.04 & 0.04 & 0.04 \\ 
Adjusted R$^{2}$ & 0.04 & 0.04 & 0.04 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table}

\hypertarget{structural-estimation}{%
\section{Structural estimation}\label{structural-estimation}}

\hypertarget{model-fit}{%
\subsection{Model fit}\label{model-fit}}

@@

I designed a setting in which I can distinguish between the forces of
dogmatism, paradigm shifts, and optimal expectations. Through a
laboratory experiment, I collected data on behavior and infer beliefs
from it. I find that the behavior of the subjects is consistent with the
self-attribution bias. In particular, I find that subjects decrease
their effort in response to bad news as well as to bad news, although
the latter decrease is smaller. This is consistent with the
self-attribution bias as well as with a change of paradigm. However, I
also observe that about a third of the subjects started with a correctly
specified belief and ended up overconfident. This is consistent only
with the self-attribution bias. Therefore, I conclude that the model
that best explains the data is that of optimal expectations.

I have small or no effects throughout on the ego-relevance of the type.
Although the motivation behind the theories does not directly apply to
the case of stereotypes, the lack of stark differences suggest that
biases arise due to an incorrect updating procedure and not necessarily
due to the self-serving motivation behind the bias. The lack of
differences in the behavior of the subjects across treatments could also
be due to self-censoring of the subjects. In particular, subjects might
be initially reluctant to express their true belief about a particular
group of people. In addition, as they learn and realize that their
beliefs might be incorrect, they might overreact to signals to try to
compensate for the initial bias. Such an updating procedure would be
consistent with the self-attribution bias for the case of sterotypes. I
remain agnostic about whether that is truly the case or whether the lack
of differences is due to other factors.

Another bias that has been studied and suggested in the literature is
the confirmation bias. In particular, it has been suggested that
subjects who are overestimate \(\theta\) in their prior, will
over-attribute successes to a high type and under-attribute them to a
low type. Symmetrically, subjects who initially underestimate \(\theta\)
will over-attribute failures to a low type and under-attribute them to a
high type. I do not find evidence of such a bias in the data. This is
consistent with what \citet{Mobius2022} find in their experiment.

I further investigate possible heterogeneity in the underlying models
being used by estimating the structural parameters of the models.
